---
title: "Exploring integer partitioning applied to constraints on biodiversity"
author: "A. J. Rominger"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The total number of species that can possibly exist is bounded above by the total number of individuals.  If there are $n$ individuals there can be at most $s = n$ species.  However, the value of $s$ is likely to be much smaller than $n$.  In fact, from a purely combinatorial perspective there are $p_k(n)$ ways of allocating $n$ individuals across $s = k$ species for abitraty $k \leq n$.  This is the realm of integer partitioning.  From a statistical mechanics perspective, the most likely number of species for a given $n$ is the number that maximizes the microstates (i.e. the value of $k$ that maximizes $p_k(n)$).  Because our world is finite, we might expect some reasonable variation about this maximum entropy $k$.  The probabilities for each $k_i$ are equal to $p_{k_i}(n) / p(n)$, where $p(n) = \sum_k p_k(n)$ is the total number of partitions of $n$ across all $k$.

## Set-up

```{r init}
library(partitions)
library(socorro)
```

## Initial exploration

Explore how the number of ways to partition an integer $n$ into $k$ parts changes across values of $n$ and $k$:

```{r npartPlot}
## max of n
N <- 100

## loop over n values, for each evaluate k in 1:n
n <- unique(round(10^seq(0, log(N, 10), length = 50)))
npart <- sapply(n, function(ni) {
    out <- sapply(1:ni, function(k) R(k, ni))
    out <- c(out, rep(NA, N - length(out)))
    return(out)
})

## plotting
par(mar = c(3.5, 3.5, 0.5, 0.5), mgp = c(2, 0.75, 0))
matplot(npart, type = 'l', log = 'xy', axes = FALSE, frame.plot = TRUE, 
        col = rainbow(length(n), end = 0.8), lty = 1, 
        xlab = 'k', ylab = 'number of partitions')
legend('topleft', 
       legend = c('n = 1', '...', 'n = 100'), pch = c(16, NA, 16), 
       col = rainbow(length(n), end = 0.8)[c(1, 1, length(n))], 
       bty = 'n')
logAxis(1)
logAxis(2)

## add the values of k that maximize number of parts
kmax <- apply(npart, 2, which.max)

points(kmax, npart[cbind(kmax, 1:length(kmax))], pch = 21, 
       bg = rainbow(length(n), end = 0.8))
```

Note the color gradient starts at $n = 1$ in red, goes through the rainbow colors, and ends at $n = 100$ in purple.  Thus each curve represents the number of partitions for a given $n$ across the full range of $k$ values $k \in \{1, \ldots, n\}$.

The filled dots show the maximum number of partitions achieved for each $n$.  We could also look at this in terms of what $k$ value maximizes the number of partitions for each $n$.  Doing so we find a power law:

```{r kmaxPlot}
plot(n, kmax, log = 'xy', axes = FALSE, frame.plot = TRUE,
     xlab = 'n', ylab = 'k giving max partitions')
logAxis(1)
logAxis(2)
mod <- lm(log(kmax) ~ log(n))
curve(exp(mod$coeff[1])*x^mod$coeff[2], add = TRUE, col = 'red')
legend('topleft', legend = paste('power =', round(mod$coeff[2], 4)), bty = 'n')
```

Note to self---check this out:
https://pdfs.semanticscholar.org/370b/06f08d729166e1bab20769a31cfad5591bc1.pdf